---
title: "Reproducible Research - Peer Assignment 1"
output: html_document
---  

### Before we start



-Download the zip file from [https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip]  
-Save the zip file to a local directory.  
-Unzip and Extract the file "activity.csv" into your current  R Studio working directory     

  
### Loading and preprocessing the data    
####1.Load the data  
The following r code will load the data as a dataframe.Make sure "activity.csv" is in your current working directory



```{r echo=TRUE}
withNA <- read.csv("activity.csv", header=TRUE)
df <- read.csv("activity.csv", header=TRUE)

```    
####2.Data Preprocessing  

Check if "steps" column is integer 
```{r echo=TRUE}

sapply(withNA,typeof)

```    
Create a dataframe with all NA data removed  
```{r echo=TRUE}

withoutNA<-withNA[complete.cases(withNA),]

```    
  
###Mean total number of steps per day
####1.Histogram of the Total Steps taken per day
```{r echo=TRUE}
totalsteps<-aggregate(withoutNA$steps ~ withoutNA$date, data=withoutNA, FUN=sum)
names(totalsteps)<-c("date","steps")
```

```{r message=FALSE}
attach(totalsteps)
```

```{r echo=TRUE}
hist(steps,main="Total Steps per Day",xlab="Total steps",col="lightgreen")
```
    
####2.Mean and Median of total steps per day
```{r echo=TRUE}
meansteps<-aggregate(withoutNA$steps ~ withoutNA$date, data=withoutNA, FUN=mean)
mediansteps<-aggregate(withoutNA$steps ~ withoutNA$date, data=withoutNA, FUN=median)
```


###Average daily activity pattern
```{r}
stepsbyinterval<-aggregate(withoutNA$steps ~ withoutNA$interval, data=withoutNA, FUN=mean)
names(stepsbyinterval)<-c("interval","steps")
```

```{r message=FALSE}
attach(stepsbyinterval)
```
####1. Time series plot of the interval and average number of steps taken

```{r echo=TRUE}
plot(interval, steps, type="l",main="Interval vs Average steps taken", xlab= "Interval", ylab= "Average steps", col="blue" , lwd=2)
```  
    
####2. 5-minute interval that contains the maximum number of steps

```{r echo=TRUE}
stepsbyinterval[order(-steps),][1,]
```
     
###Impute missing values  
####1.Total number of rows with missing values
```{r=TRUE}
nrow(withNA)-nrow(withoutNA)
```

  
####2.Strategy for filling the missing values
-Write a function to compute mean  
-Using plyr package and ddply function replace NA of steps column with the mean steps for that given interval
  
```{r echo=TRUE}
library(plyr)
meanfunc <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
```

####3.Create dataset with missing data filled in
```{r echo = TRUE}
filledNA<-(ddply(withNA, ~ interval, transform,  steps= meanfunc(steps)))
```

####4.Histogram, Mean and Median for new data with NA values filled  
```{r echo=TRUE}
totalsteps<-aggregate(filledNA$steps ~ filledNA$date, data=filledNA, FUN=sum)
names(totalsteps)<-c("date","steps")
```

```{r message=FALSE}
attach(totalsteps)
```

```{r echo=TRUE}
hist(steps,main="Total Steps per Day for data with filled NA values",xlab="Total steps",col="lightblue")
```

```{r echo=TRUE}
meansteps<-aggregate(filledNA$steps ~ filledNA$date, data=filledNA, FUN=mean)
mediansteps<-aggregate(filledNA$steps ~ filledNA$date, data=filledNA, FUN=median)
```

####5.Impact of imputing missing data
The following summaries show that imputing missing data has no effect on the actual data

```{r echo=TRUE}

summary(withoutNA)
summary(filledNA)

```

###Activity patterns - weekdays and Weekends  
####1.Creating a factor variable to indicate "Weekday" and "Weekend"

```{r}

filledNA$Daytype<-ifelse(weekdays(as.Date(filledNA$date)) %in% c('Sunday','Saturday'), 'Weekend', 'Weekday')
```
     
####2.Panel plot - time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days

```{r echo=TRUE}

library(lattice)
aggavg<-aggregate(filledNA$steps ~ filledNA$Daytype+filledNA$interval, data=filledNA, FUN=mean)
names(aggavg)<-c("Daytype","Interval","Steps")
xyplot(Steps~Interval|Daytype, group=Daytype, data=aggavg, type="l",layout=c(1,2))

```


